---
title: "PDF Scrape Tutorial"
author: "OJ Watson"
date: "04 September 2016"
output:
  html_document:
    css: knitr.css
    toc: yes
  pdf_document:
    toc: yes
---

## Problem Description

A time may arise when you find yourself having to scrape information from a list
of PDFs, which sadly are not formatted well for searching. An example of this can be seen
from looking at the the WHO website malaria country profiles:

<http://www.who.int/malaria/publications/country-profiles/en/>

This page shows a list of links to PDFs that describe country malaria profiles for
2015. These PDFs contain lots of information, but for this example let's try and
focus on the *"Reported confrmed cases:"*. 

Before we start we need to create a function to identify all the link URLs. To achieve
this and the following scraping from the pdfs we will requite a number of very useful R
packages:

## R packages required

```{r Packages, include=TRUE, message = FALSE, warning = FALSE,results='hide'}
# Install and load required packages

# Package for easy web scraping with R
install.packages("rvest",repos="http://cloud.r-project.org/")
# Useful package for string searching and manipulating
install.packages("stringr",repos="http://cloud.r-project.org/")
# Useful package for handling strange character types and escaping chracters
install.packages("qdap",repos="http://cloud.r-project.org/")
# Package for handling PDFs
install.packages("pdftools",repos="http://cloud.r-project.org/")

```

## URL List from a CSS Selector

Below is an example function that will use a provided minimal CSS selector to 
extract a vector of all the URLs that we want. There are various ways to find out 
what CSS selector you need such as using your web browser developer tools, however
a particularly neat solution is using Andrew and Kyle's brilliant SelectorGadget
<http://selectorgadget.com/> that has a fantastic chrome extension. Have a look 
at their site and see how it can be used to identify that the CSS selector *".a_z a"*
can be used to identify all the PDF links.  


```{r URL List from a CSS Selector, include=TRUE}

# Provide the url with the PDF links in and the css element
url <- "http://www.who.int/globalchange/resources/country-profiles/en/"
css <- ".a_z a"

# function to find urls of links in provided url given css
URL_Lists_From_CSS_Selector <- function(url,css){
  
  ## pull the xml files from the urls
  xml.list <- xml2::read_html(url)
  
  ## draw down the node set from the css selector
  node.set <- rvest::html_nodes(xml.list,css)
  
  ## isolate the position in contents that will corresponds to the url
  ## This is important as websites often provide internal url links rather than
  ## a complete url
  link.names <- xml2::xml_contents(node.set) 
  link.names <- rvest::html_text(link.names)
  
  ## fetch all the urls attached to the link attribute "href"
  internal.urls <- rvest::html_attr(node.set,"href")
  
  ## store these as the function output
  urls <- internal.urls
  
  ## find urls that are full urls:
  full.boolean <- unlist(lapply(strsplit(internal.urls,"/"),
                                function(x){return(x[[1]]=="http:")}))
  
  ## split the truly internal url references and the full url
  internal.split <- unlist(strsplit(internal.urls[which.min(full.boolean)],"/"))
  full.split <- unlist(strsplit(url,"/"))
  
  ## find the first match after the first case which will be a blank match
  first.common <- min(tail(match(internal.split,full.split),
                           length(match(internal.split,full.split))-1),na.rm = T)
  
  ## create the prefix url
  prefix.url <- unlist(strsplit(url,full.split[first.common]))[1]
  
  ## vector of ending urls
  end.urls <- unlist(strsplit(internal.urls[!full.boolean],
                              full.split[first.common]))[seq(2,length(internal.urls[!full.boolean])*2,2)]
  
  ## tidy the output of the truly internal urls
  urls[which(!full.boolean)] <- paste(prefix.url,full.split[first.common],end.urls,sep="")
  
  ## turn into df
  df <- data.frame(urls,row.names = link.names)
  
  return(df)
  
}

```

## Download PDFs and select text based on a common string

If we run the earlier function we will now have a vector of urls that represent
the desired PDFs:

```{r Test Download, include=TRUE, message = FALSE, warning = FALSE}

## grab pdf urls using earlier function
scraped.urls <- URL_Lists_From_CSS_Selector(url,css)

head(scraped.urls)
```

We now want to download and extract the number that appears
after *"Reported confrmed cases:"*. We will want to carry this out suppressing
errors as it is quite common for HTTP errors or PDF to text conversions
to throw errors that will not affect what we are interested in. We will 
then want to identify where our text belongs and then grab the following words.

A small warning before looking at this, it is worth mentioning that scraping is 
ususally quite messy so some of the code listed may appear unnecessary but it is
hopefully more generic in design. 

```{r Download PDF and scrape, include=TRUE, message = FALSE, warning = FALSE,results='hide'}

## Provide the string we are interested in, a location where we store the PDFs
## and what number of characters that we are interested in after the string from which
## we search for words

# String that our figure appears after
string = "Reported confirmed cases:"
# Where we store the PDFs
pdf.dir.dump = "C:/Users/Oliver/Desktop/pdfdump/"
# Here we only want the first word, but in theory you may want the 1st, 3rd, 4th etc.
words=1
# If we were storing more values then it would be nice to label our results
col.names = c("Cases")
# How many characters after the string should we search for the value we want
search.length = 200
# Boolean that will make sense later, but is set to save time on future scrapes
downloaded = FALSE

## create matrix for storing the result
mat <- matrix(nrow=length(scraped.urls$urls),ncol = length(words))
## vector that let's us know which pdfs didn't download or failed in some way
no.pdf <- rep(FALSE,length(scraped.urls$urls))

## iterate through the earlier scraped URLs
for (i in 1:length(scraped.urls$urls)){
  
  tryCatch({
    
    # Here we allow this to skip download the pdf if we have already run this once
    if(!downloaded){
      
      # Download the pdf using the scraped.urls
      download.file(url = as.character(scraped.urls$urls[i]),
                    destfile = paste(pdf.dir.dump,row.names(scraped.urls)[i],".pdf",sep=""),
                    mode = "wb")
    }
    
    # Simply turn the entire PDF into text (ugly but works)
    txt <- pdftools::pdf_text(paste(pdf.dir.dump,row.names(scraped.urls)[i],".pdf",sep=""))
    
    # Only continue if our string of interest is present
    text.present <- stringr::str_detect(txt,stringr::fixed(string))
    # Find which page it's on, again to save time
    page.no <- which(text.present)
    
    if(length(page.no)>0){
      
      # Find which positions the string of interest spans between
      pos <- stringr::str_locate(txt[page.no],stringr::fixed(string))
      
      # Split the entire text page
      tc <- strsplit(txt, "")[[page.no]]
      
      # Create one string that represents the search.length number of characters after the
      # string of interest
      string.search <-   paste(tc[(pos[2]+1):(pos[2]+search.length)],sep="",collapse = "")
      
      # Trim white space and clean and split based on white space, and then return which 
      # position of words we are interested. Here just the first
      mat[i,] <- unlist(strsplit(stringr::str_trim(qdap::clean(string.search))," "))[words]
    }
    
    # Report where this has failed so we can investigate after
  }, error=function(e){no.pdf[i] <- TRUE})
  
}

# Collate everything and name it
df <- data.frame(mat,row.names = row.names(scraped.urls))
names(df) <- col.names

```

## Executing our function

If we wrap everything above into a function - for this look at the scraper 
package within which this tutorial is also placed:

<https://github.com/OJWatson/scraper>  

```{r Execute our function, include=TRUE, message = FALSE, warning = FALSE,results='hide'}

library(devtools)
install_github("OJWatson/scraper")

t1 <- scraper::URL_PDF_Text_Pull(url = "http://www.who.int/malaria/publications/country-profiles/en/",
                       string = "Reported confirmed cases:",
                       pdf.dir.dump = "C:/Users/Oliver/Desktop/pdfdump/",
                       words=1,
                       col.names = c("Cases"),
                       downloaded = FALSE)
```

## Post processing

As we can see, the scrape has picked up the thin space that is used in these PDFs
as a thousand seperator. We can also see that some PDFs have failed, as although
the PDF exists, subtley different text is used. As such we will look at the NA
PDFs and see what the alternatives are.

```{r errors, include=TRUE, message = FALSE, warning = FALSE,}
# number of NAs
t1$Cases[1:3]
```

If we open the 2nd pdf manually we can see that the first NA is due to a different string, so we can rerun it with the alternative string.  

```{r Alternative String, include=TRUE, message = FALSE, warning = FALSE}
t2 <- scraper::URL_PDF_Text_Pull(url = "http://www.who.int/malaria/publications/country-profiles/en/",
                       string = "Total confirmed cases, 2014:",
                       pdf.dir.dump = "C:/Users/Oliver/Desktop/pdfdump/",
                       words=1,
                       col.names = c("Cases"),
                       downloaded = TRUE)

# collate these
t3 <- t1
levels(t3$Cases) <- c(levels(t3$Cases),levels(t2$Cases))
t3$Cases[is.na(t1$Cases)] <- factor(t2$Cases[is.na(t1$Cases)])

# we can now see this has no NAs
sum(is.na(t3$Cases))
```

We will then have to do some manual formatting to remove the thin space.  

```{r formatting, include=TRUE, message = FALSE, warning = FALSE}
# Remove the unicode for the thin space
t4 <- gsub(pattern = "<U+2009>",replacement = "",x=t3$Cases)

# This is almost certainly not the best way to handle this, so please if someone
# knows an alternative then let me know.

strsplit(t4[1],"")

thin.space <- strsplit(t4[1],"")[[1]][3]
# remove thin space and convert to numeric
t5 <- as.numeric(gsub(pattern=thin.space,replacement="",t4))

# Bring back to the original returned results
Result <- t1
Result$Cases <- t5

# Et voila
head(Result)
```

## Summary and further thoughts

We can see that using CSS and a variety of R packages that we can extract the 
information we want, although it is quite messy. It is also worth noting that this
is just what I have found possible from what I have discovered in R in the past 
few weeks. More recently I have found some possible alternative tools. For example,
<http://tabula.technology/> is a tool for liberating data tables locked inside PDF 
files, and has a client for R:

<https://github.com/ropenscilabs/tabulizer>

So perhaps the second half of this could be improved, however the first half shows
a neat way that rvest and CSS selectors can be very useful. Have a look at some of
the other functions in [scraper](https://github.com/OJWatson/scraper), for some 
inspiration of other uses and I hope that these can give perhaps better indications 
of what could be achieved. 

---


